Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2021-06-22 SIR_manuscript.tex
revisions to introduction

\newcommand{\fady}[1]{\comment{cyan}{Fady}{#1}}
\newcommand{\ali}[1]{\comment{magenta}{Ali}{#1}}
\newcommand{\jd}[1]{\comment{blue}{JD}{#1}}
\newcommand{\david}[1]{\comment{red}{DJDE}{#1}}
\newcommand{\bmb}[1]{\comment{red}{BMB}{#1}}
\newcommand{\todo}[1]{\comment{red}{TODO}{#1}}

@@

\title{Testing and Isolation Efficacy: Insights from a Simple Epidemic Model}

@@

\bmb{abstract needs another editing pass, skipping for now}

The effect of testing processes, including testing and test reporting, on epidemic dynamics, involving infection and recovery, can be studied at the individual level or the community level (e.g., nursing homes, long-term-care facilities, etc.).
Gaining insights to determine the sensitivity of the epidemic dynamics with respect to the testing processes will depend on underlying factors including the level of focus (individual or community), assumptions (model), and the interplay between these factors. 
In particular, fast testing and test reporting may be beneficial at the community-level, supported by many studies, as it gives a rapid assessment of the situation, identifies hot spots, and may enable rapid contact-tracing. However, the potential advantage of a slow rate of test return on the dynamics of an epidemic is real, often neglected, and needs to be quantified. At the individual level, this advantage can manifest in the following sense: individuals awaiting their test results or who have tested positive may partially or fully self-isolate, thus reducing or eliminating their potential in the transmission process.
In this paper, we investigated this individual-level effect of testing processes on the epidemic dynamics by developing a SIR-type model.
Although the model development was motivated by the \covid epidemic, the model has general epidemiological and testing structures. The realistic components of the model include \percap testing intensity, test sensitivity and specificity, rate of test return, and isolation efficacy (i.e., reduction of the transmission probability by isolating individuals). The novel component is the compartment-specific relative testing weights, which reflect the testing strategies --- surveillance, diagnosis, or control. Here, we compare two testing strategies, random vs. targeted, and concluded that increasing testing “focus” from random to targeted always decreases \Rnum. Further, the following processes always decreases $\Rnum$; increasing the isolation efficacy parameters for tested and confirmed individuals, higher testing intensity if testing is random or testing intensity is small, and a higher rate of test return if the isolation efficacy for tested-awaiting individuals is low.

@@

If testing influences behavior, then epidemic dynamics will depend on who gets tested.
The impacts of testing will depend both on testing intensity (tests performed per day) and on how strongly testing is focused on people who are infectious.
This level of focus depends in turn on the purpose and design of testing programs. 
When testing is done for the purposes of disease surveillance \citep{foddai2020base}
tests are typically assigned randomly (or using a stratified random design) across the population in order to make an unbiased assessment of population prevalence.

@@

As epidemiologists, we want to know whether an epidemic will initially grow exponentially following its introduction in a susceptible population. In mathematical epidemiology, we determine this outcome by studying the basic reproduction number $\Rnum$, defined as the expected number of secondary infections arising from a typical infective individual in a completely susceptible population \citep{dietz1993estimation}. 
In the early stages of an epidemic the number of infected individuals is expected to grow exponentially over time when $\Rnum>1$.
%% BMB left out the converse; don't think we have to be super-precise here
Although the value of $\Rnum$ cannot completely characterize the dynamics of even the simplest epidemic model \citep{shaw2021what}, it does give a simple and widely accepted index for the difficulty of control, as well as an indication of the likely final size of an epidemic \citep{ma2006generality}.

In order to understand the effect of testing processes on an epidemic dynamics, we developed a mechanistic SIR-type model with epidemic and testing components. This model provides a sensible platform to link the modeling of epidemic and testing components and study their interaction. Here, we studied the the effect of testing intensity, rate of test return and the isolation efficacy in reduction of the probability of transmission on the epidemic dynamics when different levels of testing focus (from random to highly targeted) are in place.

>>>>

\newcommand{\fady}[1]{\comment{cyan}{Fady}{#1}}
\newcommand{\ali}[1]{\comment{magenta}{Ali}{#1}}
\newcommand{\jd}[1]{\comment{blue}{JD}{#1}}
\newcommand{\david}[1]{\comment{orange}{DJDE}{#1}}
\newcommand{\bmb}[1]{\comment{red}{BMB}{#1}}
\newcommand{\todo}[1]{\comment{red}{TODO}{#1}}

@@

\title{Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model}
\author{\david{??? authors ??? affiliations ???}}

@@

\bmb{abstract needs another editing pass, skipping for now}
\david{This needs substantial tightening.}

The effect of testing processes, including testing and test reporting, on epidemic dynamics, involving infection and recovery, can be studied at the individual level or the community level (e.g., nursing homes, long-term-care facilities, etc.).
Gaining insights to determine the sensitivity of the epidemic dynamics with respect to the testing processes will depend on underlying factors including the level of focus (individual or community), assumptions (model), and the interplay between these factors. 
In particular, fast testing and test reporting may be beneficial at the community-level, supported by many studies, as it gives a rapid assessment of the situation, identifies hot spots, and may enable rapid contact-tracing. However, the potential advantage of a slow rate of test return on the dynamics of an epidemic is real, often neglected, and needs to be quantified. At the individual level, this advantage can manifest in the following sense: individuals awaiting their test results or who have tested positive may partially or fully self-isolate, thus reducing or eliminating their transmission potential.
In this paper, we investigated this individual-level effect of testing processes on the epidemic dynamics by developing a SIR-type model.
Although the model development was motivated by the \covid epidemic, the model has general epidemiological and testing structures. The realistic components of the model include \percap testing intensity, test sensitivity and specificity, rate of test return, and isolation efficacy (i.e., reduction of the transmission probability by isolating individuals). The novel component is the compartment-specific relative testing weights, which reflect the testing strategies --- surveillance, diagnosis, or control. Here, we compare two testing strategies, random vs. targeted, and concluded that increasing testing “focus” from random to targeted always decreases \Rnum. Further, the following processes always decreases $\Rnum$; increasing the isolation efficacy parameters for tested and confirmed individuals, higher testing intensity if testing is random or testing intensity is small, and a higher rate of test return if the isolation efficacy for tested-awaiting individuals is low.

@@

If testing influences behavior, then epidemic dynamics will depend on who gets tested.
The impacts of testing will depend both on testing intensity (tests performed per day) and on how strongly testing is focused on people who are infectious.
This level of focus depends in turn on the purpose and design of testing programs. 
When testing is done for the purposes of disease surveillance \citep{foddai2020base}
tests are typically conducted randomly (or using a stratified random design) across the population in order to make an unbiased assessment of population prevalence.

@@

When a new infectious disease emerges, it is important to determine whether it will grow exponentially in a susceptible population, and if so at what rate $r$.  The condition for positive exponential growth ($r>0$) is commonly expressed as $\Rnum>1$, where the basic reproduction number $\Rnum$ is the expected number of secondary infections arising from a typical infective individual in a completely susceptible population \citep{dietz1993estimation}.  Although the value of $\Rnum$ cannot completely characterize the dynamics of even the simplest epidemic model \citep{shaw2021what}, \david{well, technically it does up to rescaling time} it does give a simple and widely accepted index for the difficulty of control, as well as an indication of the likely final size of an epidemic \citep{ma2006generality}.

In order to understand the effect of testing processes on epidemic dynamics, we expanded one of the simplest mechanistic epidemic model---the standard deterministic SIR model---to include testing components. This model provides a sensible platform to link the modeling of epidemic and testing components and study their interaction. We studied the effects of testing intensity, rate of test return, and isolation efficacy, on transmission probability and epidemic dynamics when different levels of testing focus (from random to highly targeted) are in place.
